Type Error Telemetry at Scale: Pseudonymized and Stochiastic

### 2023-09-17 programming23/paper.tex
overview, size text

Both modes are opt-in. Creators have the option to make \mnonstrict{} mode
the default rather than \mnocheck{} mode.

@@

Roblox Studio collected type error telemetry in Spring 2023,
between February and April.
Every Studio session had a small random chance of generating
telemetry~(\cref{s:telemetry-records}).
The chosen sessions generated a record whenever the user switched modules and
randomly on each keystroke.
In total we collected 1.5 million telemetry records.

@@

\begin{figure}[t]
  %% TODO bigger text
  %% TODO weird artifact (|) in x-min label
  %% code/row-distribution.rkt
  \includegraphics[width=\columnwidth]{img/row-distribution.pdf}
  \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California. Shaded ranges correspond to weekends.}
  \label{f:records-per-hour}
\end{figure}

@@

\subsection{Codebase Size}
\label{s:codebase-size}

@@

There is a huge amount of variation across projects.
The largest ones have over 50k files (!) and 50k lines of code, and
there are many small ones with 1 file and/or 1 line of code.
Unsurprisingly, the numbers come with large standard deviations.
The median values are more reasonable, with roughly 3,000 lines of code
and 1,000 lines in edit ranges, but still have a large number of files: over 7,000.
The high number of files may be because creators define mavy components to
build a Roblox experience and put these components in separate files.
(\FILL{} still that's a lot of components ... any better guesses+)

\subsection{Session Size}

@@

FILL say more about why this data is interesting. What's the significance?

@@

Similar to the data on project size, there is huge variation in session
length.
The longest sessions last several days and/or contain thousands of records,
while the shortest last a few milliseconds or include just one record.
Likewise, both plots in the figure follow a power-law distribution
(especially for record count).
The median session time is approximately 14 minutes.
The median number of
records is 138.
These numbers are reasonable.

\subsection{Type Analysis Modes}

\begin{figure}[t]
  \begin{subfigure}[t]{\columnwidth}
    \begin{tabular}[t]{ll}
      \begin{tabular}[t]{r@{~~}l@{~}r}
         1,341,348 & \mnocheck{}          & [\pct{89.14}] \\
           156,883 & \mnonstrict{}        & [\pct{10.43}] \\
             6,505 & \mstrict{}           & [\pct{ 0.43}]
      \end{tabular}
      \begin{tabular}[t]{r@{~~}l@{~~}r}
           508,572 & due to module switch & [\pct{33.80}] \\
           996,164 & due to keystrokes    & [\pct{66.20}]
      \end{tabular}
    \end{tabular}
    \caption{Total telemetry records (1,504,736), analysis modes, and reason-for-sending}
    \label{f:total-records}
  \end{subfigure}

  \begin{subfigure}[t]{\columnwidth}
    \begin{tabular}[t]{ll} \\
      \begin{tabular}[t]{l@{~~}r@{~}l}
        595,137 & Type errors \\
        289,698 & in current module & [\pct{48.68}] \\
         30,924 & in edit range & [\pct{5.20}]
      \end{tabular}
      \begin{tabular}[t]{l@{~~}r@{~}l}
        72,235,735 & {Forced-strict errors} \\
        37,027,281 & in current module & [\pct{51.26}] \\
         1,111,178 & in edit range & [\pct{1.54}]
      \end{tabular}
    \end{tabular}
    \caption{Type errors and forced-strict errors across all telemetry records}
    \label{f:total-tefs}
  \end{subfigure}

  \begin{subfigure}[t]{\columnwidth}
    \begin{tabular}[t]{ll} \\
      \begin{tabular}[t]{l@{~~}r@{~}l}
        313,509 & \mnocheck{}   & [\pct{90.19}] \\
         32,902 & \mnonstrict{} & [\pct{ 9.47}] \\
            545 & \mstrict{}    & [\pct{ 0.16}] \\
            642 & mixed-mode    & [\pct{ 0.18}]
      \end{tabular}
      \begin{tabular}[t]{l@{~~}ll}
        \zerowidth{Among the mixed-mode sessions:} \\
        & 341 & contain a mode upgrade \\
        & 320 & contain a mode downgrade \\
        & 512 & contain modules with different modes
      \end{tabular}
    \end{tabular}
    \caption{Total sessions (347,598), whether and how they switch analysis mode}
    \label{f:total-sessions}
  \end{subfigure}

  \caption{FILL Type analysis overview}
  \label{f:dataset-overview}
\end{figure}

\Cref{f:dataset-overview} breaks down the dataset ...

FILL nocheck nonstrict strict, order-of-magnitude

FILL single-mode sessions, mostly

FILL details on upgrades, downgrades, then ignore these for rest of paper.
multi-mode project = found a module switch that coincides with a mode switch, which suggests that two modules have different modes;
upgrade = found a up-mode switch that does not match a module switch (quite possible to miss these);
downgrade = found a down-mode switch that does not match a module switch

wow 619 up-pairs, 583 down-pairs
=== UP
min -3 max 57 median 0 mean 2.62 stddev 6.8390518952707335
=== DOWN
min -48 max 30 median 0 mean -0.31 stddev 4.498748355271854

FILL most sessions fairly long in time (order of minutes),
short in event count (dozens).

\section{Type Errors in Edit Range}
\label{s:type-error-count}

@@

@@

@@

@@

>>>>

Both modes are opt-in. The default for a new script is \mnocheck{} by
default, but creators can change this default to \mnonstrict{}.

@@

Roblox Studio collected type error telemetry in Spring 2023,
between February and April.
Every Studio session had a small random chance of generating
telemetry~(\cref{s:telemetry-records}).
The chosen sessions generated a record whenever the user switched modules and
randomly on each keystroke.
In total, we collected 1.5 million telemetry records.
Roughly two-thirds of records are due to keystrokes.

@@

\begin{figure}[t]\centering
  %% TODO bigger text
  %% TODO weird artifact (|) in x-min label
  %% code/row-distribution.rkt
  \includegraphics[width=\columnwidth]{img/row-distribution.pdf}
  \begin{tabular}{ll}
    Reason for sending:
    &
    \begin{tabular}[t]{r@{~~}l@{~~}r}
      996,164 & due to keystrokes  & [\pct{66.20}] \\
      508,572 & due to module switch & [\pct{33.80}]
    \end{tabular}
  \end{tabular}
  \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California. Shaded ranges correspond to weekends.}
  \label{f:records-per-hour}
\end{figure}

@@

\subsection{Overall Size and Shape}

The first step toward understanding the data is to have a sense of what
is in it.
Three important characteristics are the size of the codebases in the data,
the length of the sessions we captured, and the number of type analysis
errors.
We discuss these in turn.
Codebase size and session length follow power-law distributions, with
many small items and a few very large items.
The number of type analysis errors is somewhat low~(explained in the next
section), but sufficient to proceed with futher analysis.

\paragraph{Codebase Size}
%\label{s:codebase-size}

@@

There is a huge amount of variation across projects.
The largest ones have over 50k files (!) and 50k lines of code, and
there are many small ones with 1 file and/or 1 line of code.
Unsurprisingly, the numbers come with large standard deviations.
The median values are more reasonable, with roughly 3,000 lines of code
and 1,000 lines in edit ranges, but still have a large number of files: over 7,000.
The high number of files may be because creators define mavy components to
build a Roblox experience and put these components in separate files.
(\FILL{} still that's a lot of components ... any better guesses?)

\paragraph{Session Size}
%\label{s:session-size}

@@

@@

Similar to the data on project size, there is huge variation in session
length.
The longest sessions last several days and/or contain thousands of records,
while the shortest last a few milliseconds or include just one record.
Likewise, both plots in the figure follow a power-law distribution
(especially for record count).
A typical session runs on the order of minutes and consists of a few dozen events.
The median session time is approximately 14 minutes.
The median number of
records is 138.
These numbers are reasonable.

\paragraph{Number of Type Analysis Errors}
%\label{s:count-analysis-errors}

\begin{figure}[t]
  \begin{tabular}[t]{ll} \\
    \begin{tabular}[t]{l@{~~}r@{~}l}
      595,137 & type errors \\
      289,698 & in current module & [\pct{48.68}] \\
       30,924 & in edit range & [\pct{5.20}]
    \end{tabular}
    \begin{tabular}[t]{l@{~~}r@{~}l}
      72,235,735 & {forced strict errors} \\
      37,027,281 & in current module & [\pct{51.26}] \\
       1,111,178 & in edit range & [\pct{1.54}]
    \end{tabular}
  \end{tabular}
  \caption{Type errors and forced strict errors across all telemetry records}
  \label{f:count-analysis-errors}
\end{figure}

\FILL{} need old and current in edit range

\Cref{f:count-analysis-errors} lists the number of type analysis errors
and breaks them down by location in the codebase.
In total, there are nearly 600K type errors.
These appeared in modules that opted in to typechecking with either the
\mnonstrict{} or \mstrict{} mode.
The number of forced strict errors is much larger, 72M, because this analysis
ran on every module using strict type.
(Furthermore, forced strict runs silently, so creators have no incentive to
fix these errors.)
Approximately half the analysis errors occurred in the current module,
and a small fraction overlapped with the current edit range~(\pct{5} for
type errors, \pct{1} for forced strict).

The large fraction of errors in the current module is encouraging.
It suggests that errors appear locally, as the result of edits to nearby
code (as opposed to edits in one module raising errors in another), and
that creators fix the errors before switching focus to another
module.
The low fraction of errors in the edit range suggests that errors rarely
persist through nearby edits.

\subsection{Type Analysis Modes}
\label{s:type-analysis-modes}

\begin{figure}[t]\centering
  \begin{subfigure}[t]{\columnwidth}
    \begin{tabular}[t]{ll}
      %% FILL bar chart? rectangle?
      \begin{tabular}[t]{r@{~~}l@{~}r}
         1,341,348 & \mnocheck{}          & [\pct{89.14}] \\
           156,883 & \mnonstrict{}        & [\pct{10.43}] \\
             6,505 & \mstrict{}           & [\pct{ 0.43}]
      \end{tabular}
      \begin{tabular}[t]{r@{~~}l@{~~}r}
      \end{tabular}
    \end{tabular}
    \caption{Analysis modes across the 1,504,736 records}
    \label{f:total-records}
  \end{subfigure}

  \begin{subfigure}[t]{\columnwidth}
    \begin{tabular}[t]{ll} \\
      \begin{tabular}[t]{l@{~~}r@{~}l}
        313,509 & \mnocheck{}   & [\pct{90.19}] \\
         32,902 & \mnonstrict{} & [\pct{ 9.47}] \\
            545 & \mstrict{}    & [\pct{ 0.16}] \\
            642 & mixed-mode    & [\pct{ 0.18}]
      \end{tabular}
      \begin{tabular}[t]{l@{~~}ll}
        \zerowidth{Among the mixed-mode sessions:} \\
        & 341 & contain a mode upgrade \\
        & 320 & contain a mode downgrade \\
        & 512 & contain modules with different modes
      \end{tabular}
    \end{tabular}
    \caption{Analysis mode(s) for the 347,598 sessions}
    \label{f:total-sessions}
  \end{subfigure}

  \caption{Overview of type analysis modes}
  \label{f:dataset-overview}
\end{figure}

Creators have three analysis modes to choose from.
Furthermore, they can switch between modes at any time.
According to~\cref{s:type-analysis-modes}, however, usage is
extremely skewed toward the type-free \mnocheck{} mode.
Nearly \pct{90} of all records use the \mnocheck{} mode
and \pct{90} of all sessions use \mnocheck{} exclusively.
Among the rest, roughly \pct{10} of all records use \mnonstrict{} and only a
tiny fraction (\pct{0.4}) use \mstrict{}.
The per-session data is similar: \pct{9} use \mnonstrict{} exclusively
and \pct{0.2} use \mstrict{}.

These adoption numbers are concerning.
Creators either do not know about type analysis, or they choose not
to use it in the code they are currently maintaining.
Uncovering the reason for low adaption is an important topic for future work on
Luau that has implications for the gradual typing community~(\cref{s:discussion}).

%% mostly pure, rare mixes
Most sessions (\pct{99.82}) stick to a single analysis mode.
These sessions never change the mode in the current module and never switch
focus to a module with a different analysis mode.
Among the other sessions, most (\pct{80}) switch to modules with
a different mode, about half contain at least one edit that upgrades the
current mode, and about half contain a downgrade.
There are 619 upgrades in total and 583 downgrades.
\FILL{} Ben check how often the downgrades follow the upgrades.

\paragraph{Are Upgrades Discouraging?}

%% 619 up-pairs, 583 down-pairs
%% === UP
%% min -3 max 57 median 0 mean 2.62 stddev 6.8390518952707335
%% === DOWN
%% min -48 max 30 median 0 mean -0.31 stddev 4.498748355271854

A possible explanation for the low adoption of \mnonstrict{} and
\mstrict{} mode is that upgrading to these modes leads to a large
number of type errors.
Creators might get discouraged by a high error count and revert to
\mnocheck{} mode.

The data does not support this explanation.
On average, the mode-upgrades in the data resulted in 3 additional
type errors (stddev 7, median 0).
The worst-case increase was 57 type errors, which is very high
but exceptional.
In another exception, upgrading modes removed three type errors
(presumably due to the increased precision of \mstrict{} mode
\FILL{} Alan, sounds plausible?).

Similarly, mode downgrades have only a small negative effect on the
number of errors (mean -0.3, stddev 4, median 0, max -48).
%% curious outlier: downgrade added 30 errors!
Creators apparently tend switch modes only when the code is already
well typed, or close to it.

\subsection{Type Errors in Edit Range}
\label{s:type-error-count}

@@

FILL specific type errors, 1st add/drop, 2nd popularity

@@

Findings:

\begin{itemize}
  \item
    \pct{50} of errors in current module, so probably errors appear locally (not in a distant module);
    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    \pct{90} \mnocheck{},  \pct{10} \mnonstrict{},  \pct{0.1} \mstrict{}, wow.
  \item
  \item
  \item
    DM awareness is an important and surprising gap between strict
    and forced strict, undoubtedly a usability barrier for \mstrict{}.
  \item
\end{itemize}

@@

\paragraph{Future work: Low adaption}

why so few using nonstrict and strict?
needs advertising?
needs usability?

@@

FILL these were helpful for the overview
